---
title: "Heatmaps_GDS_Final"
output: pdf_document
date: "2026-03-14"
---

## Purpose

This notebook creates three LSOA-level heat maps for the HMP Fosse Way study area:

1. Average house price by LSOA **before** the chosen event date  
2. Average house price by LSOA **after** the chosen event date  
3. Relative house price inflation by LSOA (**after vs before**)

```{r setup, message=FALSE, warning=FALSE}
library(data.table)
library(sf)
library(ggplot2)
library(dplyr)
library(scales)

knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 10,
  fig.height = 8,
  fig.align = "center"
)
```

## File paths and parameters

```{r}
base_path <- "/Users/rohankale/Documents/BSE/Geospatial DS/Final Project"

ppd_file <- file.path(base_path, "matched_ppd.csv")
distance_file <- file.path(base_path, "property_prison_distance.csv")
lsoa_file <- file.path(base_path, "fosse_way_lsoa_bbox_candidates.geojson")

event_date <- as.Date("2023-05-01")
min_sales_per_lsoa <- 3
max_distance_m <- 11000
```

## Check files exist

```{r}
file.exists(ppd_file)
file.exists(distance_file)
file.exists(lsoa_file)
```

## Load data

```{r}
ppd <- fread(ppd_file)
distance_dt <- fread(distance_file)
lsoa_sf <- st_read(lsoa_file, quiet = TRUE)
```

## Inspect columns

```{r}
colnames(ppd)
colnames(distance_dt)
colnames(lsoa_sf)
```

## Merge house-price and distance data

```{r}
houses <- merge(
  ppd,
  distance_dt,
  by = "unique_id",
  all = FALSE
)
```

## Check merged columns

```{r}
colnames(houses)
```

## Standardize the LSOA code field in the shapefile

```{r}
possible_lsoa_cols <- c("lsoa21cd", "LSOA21CD", "lsoa_code", "LSOA_CODE")
matched_lsoa_col <- intersect(possible_lsoa_cols, names(lsoa_sf))

if (length(matched_lsoa_col) == 0) {
  stop("No LSOA code column found in the shapefile.")
}

if (matched_lsoa_col[1] != "lsoa21cd") {
  lsoa_sf <- lsoa_sf %>%
    rename(lsoa21cd = all_of(matched_lsoa_col[1]))
}

lsoa_sf <- st_make_valid(lsoa_sf)
```

## Standardize key columns in the merged table

```{r}
possible_date_cols <- c("deed_date", "DEED_DATE", "date_of_transfer", "Date of Transfer")
matched_date_col <- intersect(possible_date_cols, names(houses))

if (length(matched_date_col) == 0) {
  stop("No deed-date column found after merging ppd and distance data.")
}

if (matched_date_col[1] != "deed_date") {
  setnames(houses, matched_date_col[1], "deed_date")
}

possible_price_cols <- c("price_paid", "PRICE_PAID", "sold_amount")
matched_price_col <- intersect(possible_price_cols, names(houses))

if (length(matched_price_col) == 0) {
  stop("No house-price column found after merging ppd and distance data.")
}

if (matched_price_col[1] != "price_paid") {
  setnames(houses, matched_price_col[1], "price_paid")
}

possible_lsoa_house_cols <- c("lsoa21cd", "LSOA21CD")
matched_lsoa_house_col <- intersect(possible_lsoa_house_cols, names(houses))

if (length(matched_lsoa_house_col) == 0) {
  stop("No lsoa21cd column found after merging ppd and distance data.")
}

if (matched_lsoa_house_col[1] != "lsoa21cd") {
  setnames(houses, matched_lsoa_house_col[1], "lsoa21cd")
}
```

## Basic cleaning

```{r}
houses[, deed_date := as.Date(deed_date)]
houses[, price_paid := as.numeric(price_paid)]

required_cols <- c("lsoa21cd", "price_paid", "deed_date")
missing_required <- setdiff(required_cols, names(houses))

if (length(missing_required) > 0) {
  stop(
    paste(
      "Missing required column(s):",
      paste(missing_required, collapse = ", ")
    )
  )
}

if ("distance_to_prison_m" %in% names(houses)) {
  houses <- houses[
    !is.na(distance_to_prison_m) & distance_to_prison_m <= max_distance_m
  ]
}

if ("geocode_status" %in% names(houses)) {
  houses <- houses[geocode_status == "ok"]
}

houses <- houses[
  !is.na(lsoa21cd) &
  !is.na(price_paid) &
  !is.na(deed_date)
]

houses <- houses[lsoa21cd %in% lsoa_sf$lsoa21cd]

houses[, period := ifelse(deed_date < event_date, "Before", "After")]
houses[, deed_year := format(deed_date, "%Y")]
```

## Quick summary

```{r}
summary_dt <- houses[, .(
  rows = .N,
  unique_sales = uniqueN(unique_id),
  unique_lsoas = uniqueN(lsoa21cd),
  min_price = min(price_paid, na.rm = TRUE),
  median_price = median(price_paid, na.rm = TRUE),
  max_price = max(price_paid, na.rm = TRUE)
)]

summary_dt
```

## Build average price by LSOA before the event

```{r}
before_prices <- houses[
  period == "Before",
  .(
    n_before = .N,
    avg_price_before = mean(price_paid, na.rm = TRUE)
  ),
  by = lsoa21cd
]

before_prices <- before_prices[n_before >= min_sales_per_lsoa]

before_prices
```

## Build average price by LSOA after the event

```{r}
after_prices <- houses[
  period == "After",
  .(
    n_after = .N,
    avg_price_after = mean(price_paid, na.rm = TRUE)
  ),
  by = lsoa21cd
]

after_prices <- after_prices[n_after >= min_sales_per_lsoa]

after_prices
```

## Merge before and after summaries onto the map

```{r}
map_before <- merge(lsoa_sf, before_prices, by = "lsoa21cd", all.x = TRUE)
map_after <- merge(lsoa_sf, after_prices, by = "lsoa21cd", all.x = TRUE)
```

## Heatmap 1: Average house price by LSOA before the event

```{r}
ggplot(map_before) +
  geom_sf(aes(fill = avg_price_before), color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(
    option = "magma",
    na.value = "grey90",
    labels = label_dollar(prefix = "£")
  ) +
  labs(
    title = "Average House Price by LSOA (Before Event)",
    subtitle = paste0(
      "Event date cutoff: ", event_date,
      " | LSOAs with fewer than ", min_sales_per_lsoa, " sales are omitted"
    ),
    fill = "Average price"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_blank()
  )
```

## Heatmap 2: Average house price by LSOA after the event

```{r}
ggplot(map_after) +
  geom_sf(aes(fill = avg_price_after), color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(
    option = "magma",
    na.value = "grey90",
    labels = label_dollar(prefix = "£")
  ) +
  labs(
    title = "Average House Price by LSOA (After Event)",
    subtitle = paste0(
      "Event date cutoff: ", event_date,
      " | LSOAs with fewer than ", min_sales_per_lsoa, " sales are omitted"
    ),
    fill = "Average price"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_blank()
  )
```

## Build relative house price inflation

The inflation measure used here is:

\[
\text{Inflation (\%)} = \left(\frac{\text{Average Price After}}{\text{Average Price Before}} - 1\right)\times 100
\]

```{r}
inflation_dt <- merge(
  before_prices[, .(lsoa21cd, n_before, avg_price_before)],
  after_prices[, .(lsoa21cd, n_after, avg_price_after)],
  by = "lsoa21cd",
  all = FALSE
)

inflation_dt <- inflation_dt[avg_price_before > 0]

inflation_dt[, inflation_pct := ((avg_price_after / avg_price_before) - 1) * 100]

inflation_dt
```

## Merge inflation onto the map

```{r}
map_inflation <- merge(lsoa_sf, inflation_dt, by = "lsoa21cd", all.x = TRUE)
```

## Heatmap 3: Relative house price inflation by LSOA

```{r}
ggplot(map_inflation) +
  geom_sf(aes(fill = inflation_pct), color = "white", linewidth = 0.1) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "grey90",
    labels = function(x) paste0(round(x, 1), "%")
  ) +
  labs(
    title = "Relative House Price Inflation by LSOA",
    subtitle = paste0(
      "Inflation = ((Average Price After / Average Price Before) - 1) × 100",
      " | Only LSOAs with at least ", min_sales_per_lsoa,
      " sales in both periods are shown"
    ),
    fill = "Inflation %"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_blank()
  )
```

## Optional tables for reporting

```{r}
sales_by_period <- houses[, .(
  rows = .N,
  unique_lsoas = uniqueN(lsoa21cd)
), by = period]

sales_by_period
```

```{r}
map_counts <- data.table(
  map = c(
    "Average price before",
    "Average price after",
    "Relative inflation"
  ),
  lsoas_with_values = c(
    sum(!is.na(map_before$avg_price_before)),
    sum(!is.na(map_after$avg_price_after)),
    sum(!is.na(map_inflation$inflation_pct))
  )
)

map_counts
```